Efficiency

### 2018-09-17 Second.Rnw
big commit, getting close!

However, over a number of years now the pumpjet choice has attracted sustained attention and criticism, mostly around its efficiency at low speeds, where a diesel-electric submarine would necessarily spend a great deal of its operating time.  I was commissioned by Gary Johnston to investigate this topic in detail and produce a report \parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.  Furthermore, that paper challenged the assumption that the pumpjet would be acoustically superior at low speeds, where open propellers would have no cavitation, and cause much less turbulence than the equivalent jet.

That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing, I stood by the conclusions of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

In their own evidence during the course of the day, neither of those two did choose to challenge it.  Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", whilst making clear his reluctance to make a response clear, with comments such as "I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion".  In the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.

In the August edition of Australian Defence Magazine a major article was featured on the pumpjet question \parencite{kerr2018}, referencing both my testimony in the hearing and the paper.  Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.  However, some substantial comments were provided by Dr David Kershaw, the chief of DSTG's Maritime Division, who was also Science and Technology advisor for a time on SEA 1000, the Future Submarine Project.

This paper has been commissioned by Gary Johnston of Submarines for Australia to provide a summary of the responses that have been received from official sources (in government, and Naval Group) to the earlier major report, and respond to them.  Only two substantive arguments emerge seeking to engage with the technical arguments made in the earlier paper.  The first is that the large size of the submarine now makes the pumpjet viable, where as for smaller submarines the large weight of the pumpjet would make it unviable, and that the pumpjet would be generally well-suited to smaller sumbarines as well were it not for this weight consideration.  The second is that interaction between the pumpjet and the hull obviate the evidence I have produced showing that pumpjet efficiency declines at low speeds, which are claimed to neglect considering such interactions.  This paper seeks to consider those two lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.

@@

The following gives a brief overview of the significant known public comments regarding the pumpjet that have been made by official representatives of either the Commonwealth of Australia, or Naval Group, formally known as DCNS.  Only brief quotations are selected here, with references being made to the full excerpts either in Appendices or References.

@@

This encounter shows that the viability of pumpjets was challenged some time ago.  It is noteworthy at the time that RADM Moffit makes no attempt to defend the consideration, acknowledging even that it might be easy for them to "write it off".  It is also noteworthy the tone of confidence that Senator Johnston's challenge has, driving directly for an explanation of who had the original idea, which RADM Moffit was unable or unwilling to answer.

@@

The significance of the pumpjet in Frances bid was clearly acknowledged well before the public announcement of DCNS's success.  It also provides a clear indication that the size of the submarine already appeared to be quite well-known, and was only intended to be a very minor modification from the nuclear Barracuda.  For such a small variation in displacement, it might be very reasonably expected that the length of the submarine would vary (the easiest modification), but the diameter would not need to.  These comments also display quite clearly that the pumpjet technology in France's possession is recognised as common with technology in the US and the UK.  This rules out the likelihood that some special or unique pumpjet technology, better suited just to diesel-electric submarines, had been pioneered by France alone.

@@

This comment provides clear context for the importance of the pump-jet.  Consistent with the physics outlined in Reference \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.  Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with "obsolete propeller technology" and "otherwise identical submarines".  Last accessed 4 Sept 2018 \parencite{naval2018}. Emphasis added.}
\\
\\
``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology. In a confrontation between two \textbf{otherwise identical submarines}, the one with pump jet propulsion always has the \textbf{tactical} advantage."

@@

The official explanation of the designer and builders website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a "tactical" encounter with an "otherwise identical" submarine.  As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed such as could be achieved by a diesel-electric submarine might assist in closing with.  The submarine would be at a substantial "strategic" disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.  More significantly, the speeds achieved would not allow the submarine to close with any other target from behind such as almost all surface ships, or nuclear submarines.  This statement exhibits extreme care in the wording in order to be plausibly defensible, and is aguably misleading.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)
Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?

@@

This crucial concession reveals a very different level of confidence in the pumpjet being suitable for the Australian diesel-electric submarine.  It also gives an important glimpse into the processes that both selected the winning bidder, as well as the processes for the selection of key technology components.  By revealing that the pumpjet has not yet been firmly decided upon, and that the technology choice is a product of the operational requirement, it makes it clear that the operational requirement has not yet been determined with sufficient precision to make the suitable technology choice clear to a ship-builder with ample experience in both technologies.  Moreover, it also reveals that this was also the case at the conlcusion of the CEP, and remained so well over a year after.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments. Emphasis added.}
\\
\\
``Naval Group is the manufacturer of both conventional propeller-driven submarines and larger nuclear submarines with the propulsor. The propulsor's inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology. \textbf{In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage. The propulsor was an integral part of our bid.} The Department of Defence says it intends to use the pump jet propulsion system."
\end{tcolorbox}

The verbatim quotation of the carefully worded comparison between ``otherwise identical" submarines by the global CEO more than a year after they first appeared on the Australian DCNS website demonstrate a high degree of preparation, and continued careful word selection on the topic of the pumpjet.  This statement also further reveals nuances about the CEP selection process, simultaneously declaring that the propulsor (i.e. pumpjet) was ``integral" to their bid, whilst also deferring to what the Department of Defence ``intends" do use. This exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.  It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferreed.

@@

RADM goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.  Crucially, he reveals that the 'diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet.

This is alarming for two reasons.  The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be 'tuned' to the pumpjet and the pumpjet 'tuned' to the hull.  If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.  The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumeably available to the Australian government who were looking to acquire it.

The second is that is reveals a stunning naivety in terms of basic understanding of the heirachy of engineering challenges.  It may be true that there are very many complex and challenging engineering calculations involved in building a submarine. However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is amongst the most difficult, risky or complex.  Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (from depth) is a trivially simple calculation in comparison to the complexities of, for instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation.  It was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller predecessor.

That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project.  Were it known to the competitors that their ability to achieve this trivial engineering requirement was in question, they would no-doubt be highly offended.  Any contender that was not considered competent to design a pressure hull of a new diameter should never have been invited to participate.  RADM Sammut should be called upon to clarify what he meant by ``we needed to make sure that the certifying process would match the design philosophy" because if indeed he meant that Defence had already decided that their own certifying process already precluded scaled-up diameter hulls, this represents a very bold departure from what might be considered normal risk assessment processes for major engineering projects, including those used in designing and certifying the Collins.  Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule.

Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, are futher claims regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.  This has significant bearing on the way that subsequent comments made by Defence and DCNS regarding the state of their understanding of pumpjets, and the method for their development for the Australian submarine.

@@

This comment confirms that the official policy is still to use the pumpjet, implying that the decision has indeed been made already.  This is at odds with the statement made only weeks earlier by Jean-Michel Billig.  This is also the first noteable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.  It also implies that the same pumpjet as France is fitting to the Suffren (i.e. nuclear barracuda) will not simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.  Given that the length and diameter of the hull were announced as being already decided, and seemingly very similar to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.  As pointed out in the authors testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 November 2017 - Department of Defence, in response to Question on Notice from Senator Xenophon from October 2017}]
\textit{As quoted in an ASPI Strategist blog post\parencite{davies2017}, and also by Senator Patrick in \ref{SENjohnson}, the Department of Defence argued in defence of the pump jet's efficiency.}
\\
\\
'A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.

@@

This testimony contradicts the assurance provided in an October interview with RADM Sammut.  It is clear from multiple parts of Mr Johnson's statement that the decision on whether to use the pumpjet or not, has not yet been taken, since he is ``confident... that we are going to find this pump-jet approach to be the optimum...".  Whilst acknowledging that he doesn't yet have the performance data from the nuclear Barracuda class (which is not yet in the water), he gives assurance that it will be obtained early enough to ``accommodate whateve is required", which he gives assurance is only going to be adjustments to the design of the pumpjet.  This suggests that it is still not even known whether the pump-jet that was an 'integral' part of the bid was offered in an unmodified state, or whether it was known by France that it would need to be modified from existing designs, or whether what was offered was actually only a skill-set and capability for the design of a new pumpjet.

In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an 'electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.  Nothing about this section of his testimony seems to have any reference of connection to reality.  Cross-ocean transit regularly occur at higher speeds than 17-18kt, as do cruise ships, which he mentions. These speeds are essentially unreacheable (or only extremely briefly sustainable) by a diesel electric submarine.  I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.  As I pointed out in Mr Johnson's presence in my later testimony to the Senate on a later occasion (which he could not rebut or answer in any way), it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.  Azipods have open-tipped propellers, and are not remotely related to pumpjets.  So far as I can discern, this particular part of the testimony was made-up entirely impromptu without any knowledge or care for the truth of the facts.  The degree of confidence that we can have in his assurance that the pump-jet will be the optimum approach for the submarine should be considered in light of this.

>>>>

However, over a number of years now the pumpjet choice has attracted sustained attention and criticism, mostly around its efficiency at low speeds, where a diesel-electric submarine would necessarily spend a great deal of its operating time.  I was commissioned by Gary Johnston of Submarines for Australia to investigate this topic in detail and produce a report \parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.  Furthermore, that paper challenged the assumption that the pumpjet would be acoustically superior at low speeds, where open propellers would have no cavitation, and cause much less turbulence than the equivalent jet.

That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

In their own evidence during the course of the day, neither of those two did choose to challenge it.  Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", and Senator Carr emphasising that ``it would look pretty bloody stupid after the report's handed in''. Despite these exhortations for timeliness, in the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.

In the August edition of Australian Defence Magazine a feature article on the titled `Future Submarine: Pumpjet propulsor technology choice still not well explained' \parencite{kerr2018} appeared, referencing both my testimony in the hearing and the paper.  Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.  However, some substantial comments were provided by Dr David Kershaw, the chief of DSTG's Maritime Division, who was also Science and Technology advisor for a time on SEA 1000, the Future Submarine Project.

This paper has been commissioned by Gary Johnston to provide a summary of the responses that have been received from official sources (in government, and Naval Group) to the earlier major report, and respond to them.  At the time of its original commissioning, two substantive arguments emerge seeking to engage with the technical arguments made in the earlier paper.  The first is that the large size of the submarine now makes the pumpjet viable, where as for smaller submarines the large weight of the pumpjet would make it unviable, and that the pumpjet would be generally well-suited to smaller sumbarines as well were it not for this weight consideration.  The second is that interaction between the pumpjet and the hull obviate the evidence I have produced showing that pumpjet efficiency declines at low speeds, which are claimed to neglect considering such interactions.

In early September, the Department of Defence gave a six-page written response to my paper, a full three months after the initial request. The two substantive arguments already identified were repeated, but the weight of their argument actually came to rest on another assumption.  This assumption was mentioned already by Dr Kershaw in the ADM article, but it was so completely at odds with the most basic physics that I was inclined to dismiss it as perhaps a clumsily worded statement made in an interview compounded by a little journalistic mis-quotation, rather than a substantive argument.   The re-appearance of this argument, which was spelled out and repeated multiple times in a written document that was prepared by Defence over the course of three months, well after any nominal deadline for timeliness had elapsed, dispelled all illusions that the argument was included by accident.  As such, I have also addressed it here as the first of three substantial arguments made against my earlier conclusions regarding pumpjet efficiency, which has quite significantly altered the conclusions of this paper, which seeks to consider all three lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.

This paper commences with a chronologically sequenced history of the most significant comments that have been made by public officials or Naval Group/DCNS with respect to the pumpjet, highlighting the most important implications of each in their context.  These are followed by three sections, addressing in more detail the three substantive arguments against my conclusions.  Some discussion developing the final conclusions follows, followed by appendices which include more substantial extracts or originals sources where appropriate, for reference.

@@

The following gives a brief overview of the significant known public comments regarding the pumpjet that have been made by official representatives of either the Commonwealth of Australia, or Naval Group, formally known as DCNS.  Only brief quotations are selected here, with references being made to the full excerpts either in Appendices or References.  Any emphasis in-text has been added.

@@

This encounter shows that the viability of pumpjets was challenged some time ago.  It is noteworthy at the time that RADM Moffit makes no attempt to defend the consideration, acknowledging even that it might be easy for them to ``write it off".  It is also noteworthy the tone of confidence that Senator Johnston's challenge has, driving directly for an explanation of who had the original idea, which RADM Moffit was unable or unwilling to answer.

@@

The significance of the pumpjet in Frances bid was clearly acknowledged well before the public announcement of DCNS's success.  It also provides a clear indication that the size of the submarine already appeared to be quite well-known, and was only intended to be a very minor modification from the nuclear Barracuda.  For such a small variation in displacement, it might be very reasonably expected that the length of the submarine would vary (the easiest modification), but the diameter would not need to.  These comments also display quite clearly that the pumpjet technology in France's possession is recognised as common with technology in the US and the UK.  This rules out the likelihood that some special or unique pumpjet technology, better suited just to diesel-electric submarines, had been pioneered by France alone.  This comment also directly implies that information regarding the performance of the pumpjet had already been given to the Australian  government, \textit{prior} to the conclusion of the Competitive Evaluation Process (CEP).

@@

This comment provides clear context for the importance of the pump-jet.  Consistent with the physics outlined in Reference \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.  Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear, or at least unmentioned.  As later discussed in my evidence to the Senate \ref{SENmorrison}, this is consistent with DCNS being well aware with the advantages and disadvantages of the pumpjet, whilst not making verifyably false claims.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with ``obsolete propeller technology" and ``otherwise identical submarines".  Last accessed 4 Sept 2018 \parencite{naval2018}. Emphasis added.}
\\
\\
``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology. In a confrontation between two \textbf{otherwise identical submarines}, the one with pump jet propulsion always has the \textbf{tactical} advantage."

@@

The official explanation of the designer and builder's website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a ``tactical" encounter with an ``otherwise identical" submarine.  As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed - such as could be achieved by a diesel-electric submarine - might assist in closing with.  The submarine would be at a substantial `strategic' disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.  More significantly, the speeds achieved would not allow the submarine to close with any other target from behind such as almost all surface ships, or nuclear submarines.  This statement exhibits extreme care in the wording in order to be plausibly defensible, and is aguably misleading.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)

\textbf{Journalist:} Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?

@@

This crucial concession reveals a very different level of confidence in the pumpjet being suitable for the Australian diesel-electric submarine.  It also gives an important glimpse into the processes that selected the winning bidder, and key technology components, which reveals a clear contradictions between what would have been previously foregone conclusions.  By revealing that the pumpjet has not yet been firmly decided upon, it is directly implied that France's exclusive offer to transfer the pumpjet technology could not have been the real distinguishing feature that elevated that bid above the others.  Simultaneously, the statement that technology choice is a product of the operational requirement makes it clearly implies that the operational requirement has not yet been determined with sufficient precision to make the suitable technology.  As a ship-builder with ample experience in both technologies, uniquely positioned as having fitted both pumpjets and propellers to both diesel and nuclear submarines, it is remakeable that the performance information transferred prior or during the CEP was not sufficient to allow a decision  on the technology to be made.  Moreover, it also reveals that this was not only the case at the conlcusion of the CEP, and remained so well over a year after.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments. Emphasis added.}
\\
\\
``Naval Group is the manufacturer of both conventional propeller-driven submarines and larger nuclear submarines with the propulsor. The propulsor's inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology. \textbf{In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage. The propulsor was an integral part of our bid. The Department of Defence says it intends to use the pump jet propulsion system.}"
\end{tcolorbox}

The verbatim quotation of the carefully worded comparison between ``otherwise identical" submarines by the global CEO more than a year after they first appeared on the Australian DCNS website demonstrate a high degree of preparation, and continued careful word selection on the topic of the pumpjet.  This statement also further reveals nuances about the CEP selection process, simultaneously declaring that the propulsor (i.e. pumpjet) was ``integral" to their bid, whilst also deferring to what the Department of Defence ``intends" do use. This further exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.  It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferred.

@@

RADM Sammut goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.  Crucially, he reveals that the `diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet, and aligns with emphasis given to the exact diameter of the pumpjet remaining unchanged in media report earlier in the year.\parencite{kerr2017}

This is alarming for two reasons.  The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be `tuned' to the pumpjet and the pumpjet `tuned' to the hull.  If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.  The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumeably available to the Australian government who were looking to acquire it.  Further discussion of the hull-propulsor interaction is given in Section \ref{hullpropulsor}.

The second is that is reveals a stunning naivety in terms of basic understanding of the heirachy of engineering challenges.  It may be true that there are very many complex and challenging engineering calculations involved in building a submarine. However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is amongst the most difficult, risky or complex.  Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (driven by the dive depth) is a relatively simple calculation in comparison to the complexities of many other calculations involved in submarine design. For instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation would be vastly more complex and uncertain given the unstable difficult-to-model dynamics of turbulence and cavitation involved.  In contrast, scaling up a hull-diameter was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller Swedish predecessor.  In 2006, DSTO commissioned a report which was dedicated to proposition of a different hull diameter being used for submarine used to replace Collins \parencite{joubert2006} in order to achieve lower hydrodynamic drag.  If such a development path was actually thought to impose  such significant challenges on the engineering and certification process as RADM Sammut's remarks imply, there is no way that in 2006 when the replacement was presumed to be undertaken for a smaller number of new submarines which would enter service in 2026 (as opposed to mid-2030s as currently planned) that such an undertaking would have been seriously considered.

That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project, as it would appear that he is quite alone in this belief that this really is a significant or risky hurdle in a project of this scale.  Were it known to the competitors that their ability to achieve this simple engineering requirement was in question, one might assume they would be highly offended, since any contender that was not considered competent to design a pressure hull of a new diameter would presumeably never have been invited to participate.  RADM Sammut should be called upon to clarify what he meant by ``we needed to make sure that the certifying process would match the design philosophy" because if indeed he meant that Defence had already decided that their own certifying process already precluded scaled-up diameter hulls, this represents a very bold departure from what might be considered normal risk assessment processes for major engineering projects, including those used in designing and certifying the Collins.  It would also have essentially precluded both the Germans and Japanese from the outset.  Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule, noting that the French proposal had a larger cost and longer budget than either of the other two contenders.

Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, include remarks regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.  One would assume that this meant that the French had demonstrated a willingness to supply complete and timely information on key technologies, and that RADM Sammut's exisiting experience with the French supplying such information on the pumpjet (as Sean Costello had claimed had taken place during and prior to the conclusion of the CEP) supported this view.

@@

This comment confirms that the official policy is still to use the pumpjet, implying that the decision has indeed been made already.  This is at odds with the statement made only a few weeks earlier by Jean-Michel Billig.  This is also the first noteable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.  This contradicts the notion that the same pumpjet as France is fitting to the Suffren (i.e. nuclear barracuda) will simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.  Given that the length and diameter of the hull were announced as being already decided and practically identical to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.  As pointed out in the my testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 November 2017 - Department of Defence, in response to Question on Notice from Senator Xenophon from October 2017}]
\textit{As quoted in an ASPI Strategist blog post\parencite{davies2017}, and also by Senator Patrick in \ref{SENjohnson}, the Department of Defence argued in defence of the pump jet's efficiency.}
\\
\\
`A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.

@@

This testimony contradicts the assurance provided in an October interview with RADM Sammut.  It is clear from multiple parts of Mr Johnson's statement that the decision on whether to use the pumpjet or not has not yet been taken, since he is ``confident... that we are \textit{going to find} this pump-jet approach to be the optimum...".  Whilst acknowledging that he doesn't yet have the performance data from the nuclear Barracuda class (which is not yet in the water), he claims that it will be obtained early enough to ``accommodate whatever is required", which he gives assurance is only going to be adjustments to the design of the pumpjet.  This suggests that it is still not even known whether the pump-jet that was an `integral' part of the bid was offered in an unmodified state, or whether it was known by France that it would need to be modified from existing designs, or whether what was offered was actually only a skill-set and capability for the design of a new pumpjet.

In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an `electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.  Nothing about this section of his testimony seems to have any reference of connection to reality.  Cross-ocean transits regularly occur at higher speeds than 17-18kt, including by cruise ships, which he mentions. These speeds are essentially unreacheable (or only extremely briefly sustainable) by a diesel electric submarine.  As pointed out in my evidence to Senate Economic References Committee \ref{SENmorrison} I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.  In the same session, whilst in Mr Johnson's presence, it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.  Azipods have open-tipped propellers, and are not remotely related to pumpjets.  So far as I can discern, this particular part of the testimony was made-up entirely impromptu without any knowledge or care for the truth of the matter.  The degree of confidence that we can have in his assurance that the pump-jet will be the optimum approach for the submarine should be considered in light of this.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{9 February 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick in December 2017}]
\textit{Senator Patrick inquires about the nature, cost, and conclusions of earlier research into pumpjets undertaken as part of \$214 million worth of ``de-risking" studies already undertaken, as well as the schedule and cost of subsequent research and development of the pumpjet.  Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of \$68,825.  Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.  The development and test program is not scheduled to commence until the last quarter of 2018.  Full questions and responses can be found in \ref{QON148}, \ref{QON150}, \ref{QON151}}
\\
\\
``The study concluded that pumpjets have the potential to deliver increased propulsive efficiency and lower shaft torque requirements, relative to an open propeller. This should lead to beneficial platform systems and operational impacts to increase vessel speed and/or submerged endurance.
The study developed a parametric model of pumpjet thrust and efficiency that has been used as the basis for more recent studies into pumpjet performance."
\\
\\
``Specifications for the pump jet are being developed as design of the Future Submarine proceeds. Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet. Development, testing, and assurance will be conducted by several parties including Naval Group, Direction generale de l'armement (DGA), and DSTG. Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia. Testing will be largely completed by CDR."
